---
title: "Treatment figure with SAMHSA 2022 data"
author: "Kendra Wyant"
editor_options: 
  chunk_output_type: console
format: 
  html: 
    embed-resources: true
---

```{r}
#| message: false

library(tidyverse)

theme_set(theme_classic())

devtools::source_url("https://github.com/jjcurtin/lab_support/blob/main/format_path.R?raw=true")

path_data <- ("R:/studydata/other/samhsa")

load(here::here(path_data, "NSDUH_2022.Rdata"))
```

Recode levels based on codebook
```{r}
#| message: false

d <- NSDUH_2022 |> 
  mutate(subid = QUESTID2,
         race = if_else(NEWRACE2 == 1, "Non-Hispanic White", "Other"),
         sex = if_else(irsex == 1, "man", "woman"),
         age = case_when(AGE3 == 1 ~ "12-13",
                         AGE3 == 2 ~ "14-15", 
                         AGE3 == 3 ~ "16-17",
                         AGE3 == 4 ~ "18-20",
                         AGE3 == 5 ~ "21-23",
                         AGE3 == 6 ~ "24-25",
                         AGE3 == 7 ~ "26-29",
                         AGE3 == 8 ~ "30-34",
                         AGE3 == 9 ~ "35-49",
                         AGE3 == 10 ~ "50-64",
                         AGE3 == 11 ~ "65+"),
         income = factor(IRPINC3, 
                         levels = 1:7, 
                         labels = c("< 10,000", "10,000 - 19,999", "20,000 - 29,999",
                                    "30,000 - 39,999", "40,000 - 49,999", "50,000 - 74,999",
                                    "75,000+") ),
         poverty = factor(POVERTY3, levels = 1:3, 
                          labels = c("poverty", "up to 2x federal poverty threshold",
                                     "more than 2x federal poverty threshold")),
         employment = factor(WRKSTATWK2,
                             levels = 1:9,
                             labels = c("full-time", "part-time", "student",
                                        "volunteer or has job but did not work in past week",
                                        "unemployed/looking for work", "disabled",
                                        "keeping house full-time", "retired",
                                        "not working for other reason")),
         marital = factor(irmarit, 
                          levels = 1:4,
                          labels = c("married", "widowed", "divorced or separated",
                                     "never been married")),
         education = factor(IREDUHIGHST2,
                            levels = 1:11,
                            labels = c("fifth grade or less",
                                       "sixth grade",
                                       "seventh grade",
                                       "eighth grade",
                                       "ninth grade",
                                       "tenth grade",
                                       "eleventh or twelfth grade (no diploma)",
                                       "high school diploma/GED",
                                       "some college",
                                       "Associate's degree",
                                       "college graduate or higher")),
         aud_past_year = if_else(IRPYUD5ALC == 1, "yes", "no"),
         aud_sev_past_year = factor(IRPYSEV5ALC, 
                                    levels = c(1,2,3,9),
                                    labels = c("mild", "moderate", "severe", "none")),
         aud_tx_clinician_past_year = if_else(irsutinhosp == 1 | irsutinrhab == 1 |
                                               irsutinrsmh == 1 | irsutinopl == 1 | irsutoutrhab == 1 |
                                                irsutoutmhcr == 1 | irsutoutthrp == 1 | irsutoutdoc == 1 |
                                                irsutouthosp == 1 | irsutoutschl == 1 | irsutoutopl == 1 |
                                                irsutrxalc == 1 | irsutphvid == 1 | irsutdetox == 1,
                                    "yes", "no"),
         aud_tx_clinician_carceral_past_year = if_else(irsutinhosp == 1 | irsutinrhab == 1 |
                                               irsutinrsmh == 1 | irsutinopl == 1 | irsutoutrhab == 1 |
                                                irsutoutmhcr == 1 | irsutoutthrp == 1 | irsutoutdoc == 1 |
                                                irsutouthosp == 1 | irsutoutschl == 1 | irsutoutopl == 1 |
                                                irsutrxalc == 1 | irsutphvid == 1 | irsutdetox == 1 |
                                               sutrtalcpy == 1 | sutrtdralpy == 1,
                                    "yes", "no"),
         aud_tx_any_past_year = if_else(irsutinhosp == 1 | irsutinrhab == 1 | irsutinrsmh == 1 | 
                                          irsutinopl == 1 | irsutoutrhab == 1 | irsutoutmhcr == 1 |
                                          irsutoutthrp == 1 | irsutoutdoc == 1 | irsutouthosp == 1 |
                                          irsutoutschl == 1 | irsutoutopl == 1 | irsutrxalc == 1 |
                                          irsutphvid == 1 | irsutdetox == 1  | irsutpeerspc == 1 |
                                          irsutsuppgrp == 1,
                                    "yes", "no"))
```

## AUD prevalence

11% of individuals in data set met criteria for AUD
```{r}
janitor::tabyl(d$aud_past_year)
```

Filter down to those who met criteria for AUD
```{r}
d <- d |> 
  filter(aud_past_year == "yes")
```

## Obtained AUD tx (age 12+)

### Clinican delivered
Defined as treatment for alcohol or alcohol and drugs combined in the following settings:    
- outpatient
- inpatient 
- professional tx over phone or through video
- medication tx
- detox

```{r}
janitor::tabyl(d$aud_tx_clinician_past_year)
```


### Clinician delivered (including carceral settings)

Note: can't be entirely sure about the precision of this distinction 

```{r}
janitor::tabyl(d$aud_tx_clinician_carceral_past_year)
```


### Any treatment

Includes the addition of support groups and peer support specialist or recovery coaches
```{r}
janitor::tabyl(d$aud_tx_any_past_year)
```

## Obtained AUD tx (age 18+)

### Clinican delivered

```{r}
d |> 
  filter(!age %in% c("12-13", "14-15", "16-17")) |>
  janitor::tabyl(aud_tx_clinician_past_year)
```


### Clinician delivered (including carceral settings)

```{r}
d |> 
  filter(!age %in% c("12-13", "14-15", "16-17")) |>
  janitor::tabyl(aud_tx_clinician_carceral_past_year)
```


### Any treatment

Includes the addition of support groups and peer support specialist or recovery coaches
```{r}
d |> 
  filter(!age %in% c("12-13", "14-15", "16-17")) |>
  janitor::tabyl(aud_tx_any_past_year)
```


## Figure Using clinican care for US adults

**Note: SAMHSA data for 2023 suggest this number lowers from 12% to less than 10%, but data is not available to download yet.**

```{r}
#| label: samhsa
#| out-align: right

tx_summary <- d |> 
  filter(!age %in% c("12-13", "14-15", "16-17")) |> 
  mutate(`Past year AUD` = "yes") |> 
  rename(`Receieved any treatment` = aud_tx_clinician_past_year) |>
  pivot_longer(cols = c(`Past year AUD`, `Receieved any treatment`)) |> 
  mutate(name = factor(name),
         value = if_else(value == "yes", 1, 0)) |> 
  group_by(name) |> 
  summarize(value = sum(value), 
            proportion = value/nrow(subset(d, !age %in% c("12-13", "14-15", "16-17"))), 
            .groups = "drop") |> 
  mutate(label = c("28 Million", "3.4 Million"))

unknown <- tibble(name = "Continuing care", 
                   value = pull(subset(tx_summary, name == "Receieved any treatment"), value),
                   proportion = pull(subset(tx_summary, name == "Receieved any treatment"), proportion),
            label = "?")

tx_summary |>
  ggplot(aes(x = name, y = proportion)) +
  geom_bar(stat = "identity", 
           fill = "grey",
           color = "black",
           linewidth = .7) +
  geom_bar(data = unknown,
           aes(x = name, y = proportion), 
           fill = "white", color = "black", 
           stat = "identity", 
           linetype = "dashed",
           linewidth = .7) +
  geom_text(aes(label = label), 
            vjust = -0.5,               
            color = "black",
            size = 5) +
  geom_text(data = unknown,
            aes(label = label), 
            vjust = -0.5,               
            color = "black",
            size = 5) +
  labs(x = NULL,
       y = NULL) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  theme(axis.text = element_text(size = 14),        
        axis.title = element_text(size = 16),       
        axis.line = element_line(linewidth = .7))  
```

